---
title: "A02 - Trees"
output: html_notebook
---



No need to touch the following code.  Just run it to extract the data for the assignment
```{r}
#install.packages("liver")
install.packages("liver")
library(liver)
data(churn)

# remove the state, area.code, & account.length variables, because they do not help with classification using trees
churn <- churn[,! names(churn) %in% c("state", "area.code", "account.length")]
churn

```

###1. Create a train and test dataset using the churn file retrieved from packages in R using the following code.  Use a 70% train and 30% test split of the churn data.  Omit any observations that having missing values.

The churn column is the column that contains the actual classification.  Look at the data and using the example code provided for the breast cancer example, convert the churn column to a factor so it can be used in a manner similar to the class variable in the breast cancer example.  Just convert it to a factor.  You do not need to provide lables to represent no & yes.  The breast cancer data had 2 meaning benign and 4 meaning malignant.  Thus, to make the output meaningful for the breast cancer example, the factor contained labels to represent 2 & 4.  This dataset only needs the churn column to be a factor.  No need to encode lables for no & yes.
```{r}
df <- churn
df$churn <- factor(df$churn)
df <- na.omit(df)
set.seed(1234)
index <- sample(nrow(df), 0.7 * nrow(df))
train <- df[index, ]
test  <- df[-index, ]
table(train$churn)
table(test$churn)
```

### From this point use the training data to build the model.  Use the test data to test the model by comparing the actual churn variable in the test data frame to the predicted value with the test data, using the model that you built.

### 2. Create the cp table and graph the results

```{r}
install.packages("rpart")
library(rpart)
dtree <- rpart(churn ~ ., data=train, method="class", parms=list(split="information"))
dtree$cptable
plotcp(dtree)
```

### 3. Prune the tree using the correct cp= value.

```{r}
install.packages("rattle")
library(rattle)
best_cp <- dtree$cptable[which.min(dtree$cptable[,"xerror"]), "CP"]
dtree.pruned <- prune(dtree, cp = best_cp)
fancyRpartPlot(dtree.pruned, sub = "Pruned Classification Tree")
```


### 4. Use the test dataset to determine the accuracy of the model.  Create a confusion matrix that shows Actual versus Predicted
```{r}
dtree.pred <- predict(dtree.pruned, test, type = "class")
dtree.perf <- table(Actual = test$churn, Predicted = dtree.pred)
dtree.perf
```
### 5. Discuss the results of testing dataset.  What are your thoughts on the model?

Consider the fact that unlike the breast cancer dataset, the results do not show people who were told they do not have cancer but the actually do.

This dataset will show people who left and went to another company (churn=yes) or people who stayed (churn=no).
<< type your response below >>
The model does a good job at predicting when customers will stay, but it is not as accurate at predicting who will leave. In the results, many customers who left were still predicted as staying. This means the model is better at identifying non-churners than churners. While this is not as serious as the breast cancer example, it could still cost the company since missing churners means losing customers.
<< type your response above >>

###  6. Using the train and test datasets already created, create and plot a conditional inference tree.

```{r}
install.packages("partykit")
library(partykit)
ctree.model <- ctree(churn ~ ., data = train)
plot(ctree.model, main = "Conditional Inference Tree for Churn")
```


### 7. Create a confusion matrix that shows the results of testing the model.

```{r}
ctree.pred <- predict(ctree.model, newdata = test)
ctree.perf <- table(Actual = test$churn, Predicted = ctree.pred)
ctree.perf
```

### 8. Compare the results of the classical tree to the conditional inference tree.
<< answer below this line >>
The classical decision tree gave good results but missed more customers who actually churned. The conditional inference tree performed better, correctly identifying more churners while also reducing false predictions. This shows that the conditional tree has higher accuracy and does a better job of balancing sensitivity and specificity. When comparing resuls, conditional inference tree is the stronger model for predicting churn.
<< answer above this line >> 

### 9. Use a random forest to build a model to predict churn.

```{r}
library(randomForest)
rf.model <- randomForest(churn ~ ., data = train, ntree = 500, mtry = sqrt(ncol(train)-1), importance = TRUE)
print(rf.model)
plot(rf.model, main = "Random Forest Error Rate")
```

### 10. Determine the most important variable and the least important variable.
```{r}
importance(rf.model)
varImpPlot(rf.model, main = "Variable Importance - Random Forest")
```
<< most important variable below this line >>
The most important variable in the random forest model is customer service calls. Since the customers who frequently contact customer service are more likely to be unhappy and eventually leave the company.
<< least important variable above this line >>


<< least important variable below this line >>
The least important variable in the model is phone number. This is just an identifier and has no relevance on whether a customer stays or churns.
<< least important variable above this line >>


### 11. Test the random forest model to determine whether it was good model.
```{r}
rf.pred <- predict(rf.model, newdata = test)
rf.perf <- table(Actual = test$churn, Predicted = rf.pred)
rf.perf
rf.accuracy <- sum(diag(rf.perf)) / sum(rf.perf)
rf.accuracy
```


### 12.  Which of the models performed the best (classical, conditional inference, randomForest? Explain.
<< answer below this line >>
The randomForest model performed the best out of the three. It showed the most accurate and made the least mistakes when predicting whether customers would leave or stay. The conditional inference tree did better than the classical tree, but it was not as strong as the random forest. Overall, the random forest is the best choice for this dataset because it gave the most reliable results. I would rate them from best to least as: randomForest, conditional, classical and inference.
<< answer above this line >>


